\documentclass[11pt]{article}
\usepackage{changepage}
\usepackage{hyperref}
\usepackage{amssymb}
\usepackage{amsmath}
\begin{document}

\begin{center}
    \vspace{0.5cm}
    Investigating the Acceleration Vector and Geodesics on $S^2$\\
    MA3101 Project\\[24pt]
    \LARGE

    \Large
    \textbf{Robert Davidson}\\[6pt]
    \small
    BSc Mathematics and Computer Science \\ r.davidson1@universityofgalway.ie\\[6pt]
    Date: \today\\[12pt]
\end{center}

\vspace{1.5cm}

\subsection*{Why this project?}
In my Euclidean and Non Euclidean Geometry lectures, we learned that if the acceleration vector of a curve is always normal to the tangent plane, then the curve is a geodesic\\[0.5ex]
On the sphere, $S^2$, every geodesic is a great circle, which is circle formed a plane through the sphere's center. \\[0.5ex]
I was really intrigued by this result, and I wanted to find a way to visualise it, which led me to build an interactive visualization, which is avilable at \href{https://robertdavidson1.github.io/MA3101Project/}{this website} hosted on GitHub.
This brief writeup covers the mathematics I did to both understand why the statement is true, and to create the website. 

\subsection*{The Motivation}
In order to explore the statement, that all geodesics on $S^2$ have acceleration vectors normal to the tangent plane, we first need to define the geometry of $S^2$ and what it means for a vector to be normal to the tangent plane. \\[0.5ex] We then move onto to defining the plane so we can define how it intersects the sphere. \\[0.5ex]
Once we have this plane, we can define a circle on this plane, and parameterize it, and take the derivatives in order to get the velocity and acceleration vectors. \\[0.5ex]
Finally, we tie it all together to show that if the acceleration vector of a curve is always normal to the tangent plane, then the curve is a geodesic. 

\newpage

\subsection*{The 2-Sphere}
We define the 2-sphere, $S^2$, as the set of all points in $\mathbb{R}^3$ such that:
$$x^2 + y^2 + z^2 = 1$$
This represents a sphere of radius 1 centered at the origin. \\
The normal vector at a point $\mathbf{s} = (x, y, z)$ on the 2-sphere is the vector perpendicular to the tangent plane at that point. \\
It is given by the gradient of the function $f(x,y,z) = x^2 + y^2 + z^2 - 1$:
$$\mathbf{n}_{S^2} = \nabla f(\mathbf{s}) = (2x, 2y, 2z) = 2\textbf{s}$$
So, the normal vector is just a scalar multiple of the position vector. 
\subsection*{The Plane}
A plane $P$ can be written as:
$$ax + by + cz = d$$
Where $\textbf{n} = (a,b,c)$ is the normal vector to the plane, so the equation becomes:
$$P : \textbf{n} \cdot (x,y,z) = d$$
We normalize $\textbf{n}$ so it represents direction only:
$$\hat{\textbf{n}} = \frac{\textbf{n}}{\|\textbf{n}\|}$$
We can then write the plane equation as:
$$P : \hat{\textbf{n}} \cdot (x,y,z) = \frac{d}{\|\textbf{n}\|}$$
We let $h = \frac{d}{\|\textbf{n}\|}$ be the distance from the origin to the plane.
So the plane equation becomes:
$$P : \hat{\textbf{n}} \cdot (x,y,z) = h$$
We note that the plane intersects the sphere when $|h| \leq 1$.
\newpage
\subsection*{The Intersection of a Plane and the 2-Sphere}
We define a line $L(t)$ a line from the origin in the direction of the planes normal vector: $L(t) = t \hat{\textbf{n}}$. \\
The line intersects the plane when $t= h$, so the closest point on the plane to the origin is:
$$\mathbf{C} = h \hat{\mathbf{n}}
$$
The plane cuts the sphere in a circle centered at $\mathbf{C}$.
If we let:
\noindent



    \vspace{0pt}
\begin{itemize}
    \item Q : A point on the circumference of the circle formed by the intersection of the plane and the sphere
    \item C : The center of the circle
    \item O : The origin
\end{itemize}
We form a $\triangle OQC$ with sides:
\begin{itemize}
    \item $OC = h$ : Distance from origin to center of circle
    \item $OP = R$ : Distance from origin to point on circumference
    \item $CP = r$ : Distance from center of circle to point on circumference of planes cut
\end{itemize}
By the Pythagorean theorem, we have:
\begin{align*}
    R^2 & = h^2 + r^2                                               \\
    r   & = \sqrt{R^2 - h^2}                                        \\
    r   & = \sqrt{1 - h^2} \quad \text{Since radius of sphere is 1}
\end{align*}
So the circle on the sphere, cut by plane has center, $\mathbf{C} = (0,0,h)$ and radius $r = \sqrt{1 - h^2}$. \\
For simplicty, in the website, we let $\hat{\textbf{n}} = (0,0,1)$, which also helped in choosing vectors $\mathbf{u}$ and $\mathbf{v}$ later, then we get
\begin{align*}
     & P : (0,0,1) \cdot (x,y,z) = h    \\
     & P : z = h                        \\[1ex]
     & \mathbf{C} = h \hat{\textbf{n}} = (0,0,h) \\
     & r = \sqrt{1 - h^2}
\end{align*}
\subsection*{Parameterization of the Circle on the Plane}
We choose two vectors, \textbf{u} and \textbf{v}, that are on the plane, that is:
$$\mathbf{u}, \textbf{v} \perp \hat{\textbf{n}}$$
If we enforce $$\mid \mid \mathbf{u} \mid \mid = \mid \mid \textbf{v} \mid \mid = 1 \quad \text{and} \quad \mathbf{u} \perp \textbf{v}$$
we can define a circle on P:
$$\textbf{p}(t) = \cos(t) \mathbf{u} + \sin(t) \textbf{v} \quad t \in [0, 2\pi]$$
More precisely, $\textbf{p}(t)$ is a position vector from the origin to a point on the circle on the plane. \\
Recognising that we let $\hat{\textbf{n}} = \hat{k} = (0,0,1$), and using the fact:
$$\hat{i} \perp \hat{j} \perp \hat{k} \quad \text{and} \quad \mid \mid \hat{i} \mid \mid = \mid \mid \hat{j} \mid \mid = \mid \mid \hat{k} \mid \mid = 1$$
We can let $\textbf{u} = \hat{i}$ and $\textbf{v} = \hat{j}$, then we get:
$$\textbf{p}(t) = \cos(t) \hat{i} + \sin(t) \hat{j}\quad t \in [0, 2\pi]$$
Finally, scaling to radius $r$ and moving to center $\mathbf{c}$ we get:
$$\textbf{p}(t) = \mathbf{C} + r \left[\cos(t) \hat{i} + \sin(t) \hat{j}\right]$$
\subsection*{Velocity and Acceleration Vectors}
We can take the first and second derivatives of $\textbf{p}(t)$ to get the velocity and acceleration vectors:
\begin{align*}
    \mathbf{v}(t) & = \frac{d}{dt} \left[\textbf{p}(t)\right] =  r\left[-\sin(t) \hat{i} + \cos(t) \hat{j} \right]\\                                                               
          \mathbf{a}(t) &= -r[\cos(t)\hat{i} + \sin(t)\hat{j}] 
\end{align*}
Rewriting \textbf{p}(t), we get:
$$r[\cos(t)\,\hat{i} + \sin(t)\,\hat{j}] = \textbf{p}(t) - \mathbf{C}$$
So, we can rewrite $\mathbf{a}(t)$ as:
$$\mathbf{a}(t) = \mathbf{C} - \textbf{\textbf{p}(t)}$$
\newpage
\subsection*{Tieing it all together}
We see that if $h = 0$, that is the distance from the origin to the plane is 0, the plane gives us a circle on $S^2$ with center at the origin and radius 1, which is the definition of a great circle.
\begin{align*}
    \mathbf{C} &= h \hat{\textbf{n}} = 0 \hat{\textbf{n}} = 0  \\[1ex]
    r &= \sqrt{1- h^2} = \sqrt{1- 0^2} = 1      \\
\end{align*}
The aceleration vector also simplifies to:
\begin{align*}
    \mathbf{a}(t) & = \mathbf{C} - \textbf{p}(t) \\
    \mathbf{a}(t) & = 0 - \textbf{p}(t) \\
    \mathbf{a}(t) & = - \textbf{p}(t)
\end{align*}
So when $h = 0$,that is, when the plane cuts the 2-Sphere through the origin, we get a circle centered at the origin with radius 1, this by definition is a great circle. \\
The acceleration vector around this circle is just a scalar multiple of the position vector, which is the same as the sphere's normal vector. \\
So, when     \\[1ex]
Therefore, in order for the acceleration vector to be normal to the tangent plane, the curve must be a great circle, and thus a geodesic.
\pagebreak
\section*{Acknowledgements}
I used ChatGPT to help verify the algebraic derivations and ensure the mathematical reasoning in this write-up was correct.

\end{document}